------------------------------------------------------------------------

---
title: "illumination publication figure"
---

```{r}

library(colorEvoHelpers)
library(dplyr)
library(tidyr)
library(phyr)
library(dplyr)
library(car)
library(performance)
library(MuMIn)
library(buildmer)
library(glmmTMB)
library(splines)
library(sjPlot)
library(ggeffects)
library(ggplot2)
library(brms)
library(sjPlot)

# load your main target
d <- readRDS("_targets/objects/d_sp_cell")
```

Sample extreme cases of positive and negative skewness

```{r}

# Find species-cells with extreme skewness values
extreme_cases <- d %>%
  filter(abs(lightness_skew) > 1.5,  # Very skewed
         n >= 20) %>%                # Good sample size
  arrange(lightness_skew) %>%
  slice(c(1:3, (n()-2):n()))         # Take 3 most negative, 3 most positive

extreme_cases %>%
  unnest(lightness_values) %>%
  ggplot(aes(x = lightness_values)) +
  geom_histogram(bins = 20, alpha = 0.7, fill = "lightblue") +
  facet_wrap(~paste0(species, " | ", cell, 
                     "\nSkew: ", round(lightness_skew, 2)))

```

Visualize more species regimes

```{r}

# Sample from different lighting bias categories
strategic_sample <- d %>%
  mutate(bias_category = case_when(
    lightness_skew < -1 ~ "Underexposed",
    lightness_skew > 1 ~ "Overexposed", 
    TRUE ~ "Balanced"
  )) %>%
  group_by(bias_category) %>%
  slice_sample(n = 4) %>%  # 4 from each category
  ungroup()

strategic_sample %>%
  unnest(lightness_values) %>%
  ggplot(aes(x = lightness_values, fill = bias_category)) +
  geom_histogram(bins = 15, alpha = 0.7) +
  facet_wrap(~paste0(species, "\n", bias_category, 
                     "\nSkew: ", round(lightness_skew, 2))) +
  scale_fill_manual(values = c("blue", "gray", "red")) +
  theme(legend.position = "none")

colnames(d)
```

Simulate ideal curves describing the general effect of lighting noise

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(moments)

# tunable parameters
n_images <- 500         # Total samples for distribution calculation
n_patches_display <- 50 # Number of color patches to display (subset of n_images)
seed_val <- 123
# balanced distribution parameters (normal distribution)
balanced_mean <- 50
balanced_sd <- 15
# shared gamma parameters for mirrored distributions
gamma_shape <- 1.5      # Controls tail length (smaller = longer tail)
gamma_rate <- 0.06      # Controls spread (smaller = wider spread)
# underexposed parameters (gamma flipped for negative skew)
under_max_val <- 75     # Maximum before flipping
under_shift <- -10      # Fine-tune brightness after flipping
# overexposed parameters (gamma direct for positive skew)
over_min_val <- 15      # Minimum before shifting
over_shift <- 20      # Fine-tune brightness after shifting
# hard truncation limits
min_lightness <- 0
max_lightness <- 100

# Generate simulated lightness distributions
set.seed(seed_val)

# Balanced lighting (symmetric normal distribution)
balanced <- rnorm(n_images, mean = balanced_mean, sd = balanced_sd)
balanced <- pmax(min_lightness, pmin(max_lightness, balanced))

# Underexposed (negative skew - peak bright, long tail toward dark)
# Generate gamma, then flip to create negative skew
under_raw <- rgamma(n_images, shape = gamma_shape, rate = gamma_rate)
# Flip the distribution: high gamma values become low lightness values
underexposed <- under_max_val - under_raw + under_shift
underexposed <- pmax(min_lightness, pmin(max_lightness, underexposed))

# Overexposed (positive skew - peak dark, long tail toward bright)  
# Generate gamma directly for positive skew (same parameters as underexposed)
over_raw <- rgamma(n_images, shape = gamma_shape, rate = gamma_rate)
overexposed <- over_min_val + over_raw + over_shift
overexposed <- pmax(min_lightness, pmin(max_lightness, overexposed))

# Combine into data frame
lighting_data <- data.frame(
  lightness = c(balanced, underexposed, overexposed),
  condition = rep(c("Balanced", "Underexposed", "Overexposed"), each = n_images),
  image_id = 1:(3 * n_images)
)

# Calculate skewness for each condition
skew_stats <- lighting_data %>%
  group_by(condition) %>%
  summarise(
    skewness = round(moments::skewness(lightness), 2),
    mean_light = round(mean(lightness), 1),
    median_light = round(median(lightness), 1),
    q25 = round(quantile(lightness, 0.25), 1),
    q75 = round(quantile(lightness, 0.75), 1),
    .groups = 'drop'
  )

print("Distribution statistics:")
print(skew_stats)

# Create color patches based on lightness values
create_color_patches <- function(lightness_values, title, n_cols = 10) {
  # Sample a subset for display
  if (length(lightness_values) > n_patches_display) {
    display_values <- sample(lightness_values, n_patches_display)
  } else {
    display_values <- lightness_values
  }
  
  n_patches <- length(display_values)
  n_rows <- ceiling(n_patches / n_cols)
  
  # Shuffle the selected lightness values for random arrangement
  shuffled_lightness <- sample(display_values)
  
  # Create grid positions
  patch_data <- data.frame(
    x = rep(1:n_cols, length.out = n_patches),
    y = rep(n_rows:1, each = n_cols)[1:n_patches],
    lightness = shuffled_lightness
  )
  
  # Create the patch plot
  ggplot(patch_data, aes(x = x, y = y, fill = lightness)) +
    geom_tile(color = "white", size = 0.1) +
    scale_fill_gradient(low = "black", high = "white", 
                       name = "Lightness", limits = c(0, 100)) +
    coord_equal() +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12),
      legend.position = "none"
    ) +
    labs(title = paste0(title, "\n(", n_patches, "/", length(lightness_values), " images shown)"))
}

# Create distribution plots with enhanced tail visualization
create_distribution_plot <- function(data_subset, title, skew_val) {
  mean_val <- mean(data_subset$lightness)
  
  ggplot(data_subset, aes(x = lightness)) +
    geom_histogram(aes(y = ..density..), bins = 12, alpha = 0.7, 
                   fill = "grey", color = "black", size = 0.3) +
    #geom_density(color = "red", size = 1.2, adjust = 1.2) +
    geom_vline(aes(xintercept = mean_val), 
               color = "blue", linetype = "dashed", size = 1) +
    #geom_vline(aes(xintercept = median(lightness)), 
    #           color = "orange", linetype = "dotted", size = 1) +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(
      title = paste0("Skewness = ", skew_val),
      x = "",
      y = ""
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      panel.grid.minor = element_blank()
    )
}

# Generate plots for each condition
conditions <- c("Underexposed", "Balanced", "Overexposed")
patch_plots <- list()
dist_plots <- list()

for (i in 1:3) {
  condition <- conditions[i]
  condition_data <- lighting_data[lighting_data$condition == condition, ]
  skew_val <- skew_stats$skewness[skew_stats$condition == condition]
  
  # Create patch plot
  patch_plots[[i]] <- create_color_patches(
    condition_data$lightness,
    condition,
    n_cols = 10
  )
  
  # Create distribution plot
  dist_plots[[i]] <- create_distribution_plot(
    condition_data, 
    condition, 
    skew_val
  )
}

# Combine all plots
final_plot <- (patch_plots[[1]] | patch_plots[[2]] | patch_plots[[3]]) /
              (dist_plots[[1]] | dist_plots[[2]] | dist_plots[[3]])

# Add overall title
final_plot <- final_plot + 
  plot_annotation(
    title = "Lighting Bias Simulation Using Gamma Distributions",
    subtitle = "Color patches show simulated image lightness | Histograms show distribution shape",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 12, hjust = 0.5))
  )

# Display the plot
print(final_plot)
```

Visualize extremely skewed species-cells in the same way

```{r}

library(ggplot2)
library(dplyr)
library(patchwork)
library(moments)
library(tidyr)

# === SELECTION PARAMETERS ===
# Adjust these to control which species-cells to show
min_sample_size <- 25        # Minimum images per species-cell
extreme_skew_threshold <- 1.0  # Minimum |skewness| for extreme cases
n_examples_per_category <- 2   # 2 examples per lighting category
patch_cols <- 10           # Number of columns in color patch grid
max_patches_per_cell <- 30   # Maximum number of color patches to show per species-cell

# Set seed for reproducible selection
set.seed(456)
# === END PARAMETERS ===

# Select representative examples from your real data
real_examples <- d %>%
  filter(n >= min_sample_size) %>%  # Good sample sizes only
  mutate(
    bias_category = case_when(
      lightness_skew < -0.5 ~ "Underexposed",
      lightness_skew > 0.5 ~ "Overexposed", 
      abs(lightness_skew) < 0.3 ~ "Balanced",
      TRUE ~ "Moderate"
    ),
    abs_skew = abs(lightness_skew)
  ) %>%
  filter(bias_category != "Moderate") %>%  # Only clear examples
  group_by(bias_category) %>%
  # Select the most extreme examples within each category
  slice_max(abs_skew, n = n_examples_per_category, with_ties = FALSE) %>%
  ungroup() %>%
  # Reorder: Underexposed, Balanced, Overexposed
  mutate(bias_category = factor(bias_category, 
                                 levels = c("Underexposed", "Balanced", "Overexposed"))) %>%
  arrange(bias_category, desc(abs_skew))  # Order by category, then by extremeness

# Print selected examples info
cat("Selected examples:\n")
print(real_examples %>% 
  select(species, cell, bias_category, lightness_skew, mean_lightness, n))

# Create distribution plots for real data
create_real_distribution_plot <- function(lightness_values, species_name, 
                                        skew_val, bias_category) {
  
  # Create data frame for plotting
  plot_data <- data.frame(lightness = lightness_values)
  
  # Color based on bias category
  fill_color <- case_when(
    bias_category == "Underexposed" ~ "lightgray",
    bias_category == "Overexposed" ~ "lightgray", 
    bias_category == "Balanced" ~ "lightgray",
    TRUE ~ "lightyellow"
  )
  
  ggplot(plot_data, aes(x = lightness)) +
    geom_histogram(aes(y = ..density..), bins = 15, alpha = 0.7, 
                   fill = fill_color, color = "black") +
    geom_vline(aes(xintercept = mean(lightness)), 
               color = "blue", linetype = "dashed", size = 1) +
    scale_x_continuous(limits = c(0, 100)) +
    scale_y_continuous(labels = NULL) +
    labs(
      x = "",
      y = ""
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 10),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}

# Generate plots for each selected example
n_examples <- nrow(real_examples)
patch_plots <- list()
dist_plots <- list()

for (i in 1:n_examples) {
  example <- real_examples[i, ]
  lightness_vals <- example$lightness_values[[1]]
  
  # Create patch plot
  patch_plots[[i]] <- create_real_color_patches(
    lightness_vals,
    example$species,
    example$cell,
    example$lightness_skew
  )
  
  # Create distribution plot  
  dist_plots[[i]] <- create_real_distribution_plot(
    lightness_vals,
    example$species,
    example$lightness_skew,
    example$bias_category
  )
}

# Create a single row layout: 6 patches on top, 6 distributions below
final_plot <- (patch_plots[[1]] | patch_plots[[2]] | patch_plots[[3]] | 
               patch_plots[[4]] | patch_plots[[5]] | patch_plots[[6]]) /
              (dist_plots[[1]] | dist_plots[[2]] | dist_plots[[3]] | 
               dist_plots[[4]] | dist_plots[[5]] | dist_plots[[6]])

# Add overall title
final_plot <- final_plot + 
  plot_annotation(
    title = "",
    subtitle = "",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 12, hjust = 0.5))
  )

# Display the plot
print(final_plot)

# === PRINT DETAILED IDS ORDERED BY LIGHTNESS ===
cat("\n\n")
cat(rep("=", 120), "\n", sep="")
cat("IDS ORDERED BY LIGHTNESS FOR SELECTED SPECIES-CELLS\n")
cat(rep("=", 120), "\n", sep="")

# Collect all IDs for later use
all_ids <- c()

for (i in 1:nrow(real_examples)) {
  example <- real_examples[i, ]
  lightness_vals <- example$lightness_values[[1]]
  ids_vals <- example$ids[[1]]
  urls_vals <- example$urls[[1]]
  
  # Create data frame and sort by lightness
  sorted_data <- data.frame(
    lightness = lightness_vals,
    id = ids_vals,
    url = urls_vals
  ) %>%
    arrange(lightness)
  
  # Append sorted IDs to the combined list
  all_ids <- c(all_ids, sorted_data$id)
  
  # Print header for this species-cell
  cat("\n", rep("-", 120), "\n", sep="")
  cat(sprintf("SPECIES: %s | CELL: %s | Category: %s | Skew: %.3f\n", 
              example$species, 
              example$cell,
              as.character(example$bias_category),
              example$lightness_skew))
  cat(rep("-", 120), "\n", sep="")
  cat(sprintf("Total images: %d\n\n", nrow(sorted_data)))
  
  # Print each URL with its lightness value and ID
  for (j in 1:nrow(sorted_data)) {
    cat(sprintf("[%3d] L* = %6.2f | %s | %s\n", 
                j, 
                sorted_data$lightness[j],
                sorted_data$id[j],
                sorted_data$url[j]))
  }
}

cat("\n", rep("=", 120), "\n", sep="")

# === PRINT ALL IDS AS PYTHON LIST ===
cat("\n\n")
cat(rep("=", 120), "\n", sep="")
cat("ALL IDS AS PYTHON LIST\n")
cat(rep("=", 120), "\n", sep="")

# Print as Python list with quotes around each ID
cat("\nids = [")
cat(paste0('"', all_ids, '"', collapse = ", "))
cat("]\n")

cat(sprintf("\n# Total IDs: %d\n", length(all_ids)))
cat(rep("=", 120), "\n", sep="")
```

Get the URLs of images in the tails, to manually add to the plot

```{r}

# Function to extract URLs from the tails of lightness distributions
extract_tail_urls <- function(species_name, cell_id, lightness_values, urls, 
                               tail_percentile = 0.1, n_urls_per_tail = 5) {
  
  # Calculate percentile thresholds
  lower_threshold <- quantile(lightness_values, tail_percentile)
  upper_threshold <- quantile(lightness_values, 1 - tail_percentile)
  
  # Find indices of images in each tail
  dark_tail_indices <- which(lightness_values <= lower_threshold)
  bright_tail_indices <- which(lightness_values >= upper_threshold)
  
  # Sample URLs from each tail (in case there are many)
  dark_urls <- if(length(dark_tail_indices) > 0) {
    sample_indices <- if(length(dark_tail_indices) > n_urls_per_tail) {
      sample(dark_tail_indices, n_urls_per_tail)
    } else {
      dark_tail_indices
    }
    urls[sample_indices]
  } else {
    character(0)
  }
  
  bright_urls <- if(length(bright_tail_indices) > 0) {
    sample_indices <- if(length(bright_tail_indices) > n_urls_per_tail) {
      sample(bright_tail_indices, n_urls_per_tail)
    } else {
      bright_tail_indices
    }
    urls[sample_indices]
  } else {
    character(0)
  }
  
  # Return structured results
  list(
    species = species_name,
    cell = cell_id,
    dark_tail = list(
      threshold = round(lower_threshold, 1),
      n_images = length(dark_tail_indices),
      urls = dark_urls,
      lightness_values = lightness_values[if(length(dark_urls) > 0) match(dark_urls, urls) else integer(0)]
    ),
    bright_tail = list(
      threshold = round(upper_threshold, 1), 
      n_images = length(bright_tail_indices),
      urls = bright_urls,
      lightness_values = lightness_values[if(length(bright_urls) > 0) match(bright_urls, urls) else integer(0)]
    )
  )
}

# Function to print tail URLs in a readable format
print_tail_urls <- function(tail_results) {
  cat("\n" , rep("=", 80), "\n", sep="")
  cat("SPECIES:", tail_results$species, "| CELL:", tail_results$cell, "\n")
  cat(rep("=", 80), "\n", sep="")
  
  # Dark tail (underexposed images)
  cat("\nDARK TAIL (Underexposed):\n")
  cat("Threshold: ≤", tail_results$dark_tail$threshold, "lightness units\n")
  cat("Total images in dark tail:", tail_results$dark_tail$n_images, "\n")
  cat("Showing", length(tail_results$dark_tail$urls), "URLs:\n\n")
  
  if(length(tail_results$dark_tail$urls) > 0) {
    for(i in 1:length(tail_results$dark_tail$urls)) {
      cat(sprintf("  [Dark %d] L*=%5.1f | %s\n", 
                  i, 
                  tail_results$dark_tail$lightness_values[i],
                  tail_results$dark_tail$urls[i]))
    }
  } else {
    cat("  No URLs in dark tail\n")
  }
  
  # Bright tail (overexposed images)
  cat("\nBRIGHT TAIL (Overexposed):\n")
  cat("Threshold: ≥", tail_results$bright_tail$threshold, "lightness units\n") 
  cat("Total images in bright tail:", tail_results$bright_tail$n_images, "\n")
  cat("Showing", length(tail_results$bright_tail$urls), "URLs:\n\n")
  
  if(length(tail_results$bright_tail$urls) > 0) {
    for(i in 1:length(tail_results$bright_tail$urls)) {
      cat(sprintf("  [Bright %d] L*=%5.1f | %s\n", 
                  i, 
                  tail_results$bright_tail$lightness_values[i], 
                  tail_results$bright_tail$urls[i]))
    }
  } else {
    cat("  No URLs in bright tail\n")
  }
  
  cat("\n")
}

# Function to open URLs from tails in browser
open_tail_urls <- function(tail_results, open_dark = TRUE, open_bright = TRUE, delay = 0.5) {
  
  cat("Opening URLs for", tail_results$species, "cell", tail_results$cell, "\n")
  
  if(open_dark && length(tail_results$dark_tail$urls) > 0) {
    cat("Opening", length(tail_results$dark_tail$urls), "dark tail URLs...\n")
    for(url in tail_results$dark_tail$urls) {
      browseURL(url)
      Sys.sleep(delay)
    }
  }
  
  if(open_bright && length(tail_results$bright_tail$urls) > 0) {
    cat("Opening", length(tail_results$bright_tail$urls), "bright tail URLs...\n") 
    for(url in tail_results$bright_tail$urls) {
      browseURL(url)
      Sys.sleep(delay)
    }
  }
}

# === MAIN EXECUTION ===
# Apply to your selected examples (assuming 'real_examples' exists from previous code)

# Parameters
tail_percentile <- 0.15  # Look at most extreme 15% on each end
n_urls_per_tail <- 10     # Show 3 URLs from each tail

# Process each selected species-cell
tail_results_list <- list()

for(i in 1:nrow(real_examples)) {
  example <- real_examples[i, ]
  
  # Extract the data
  lightness_vals <- example$lightness_values[[1]]
  urls <- example$urls[[1]]
  
  # Get tail URLs
  tail_results <- extract_tail_urls(
    species_name = example$species,
    cell_id = example$cell,
    lightness_values = lightness_vals,
    urls = urls,
    tail_percentile = tail_percentile,
    n_urls_per_tail = n_urls_per_tail
  )
  
  # Store results
  tail_results_list[[i]] <- tail_results
  
  # Print results
  print_tail_urls(tail_results)
}

# === OPTIONAL: Open URLs in browser ===
# Uncomment the following lines to open URLs in your browser:

# # Open dark tail URLs only (underexposed images)
# lapply(tail_results_list, function(x) open_tail_urls(x, open_dark = TRUE, open_bright = FALSE))

# # Open bright tail URLs only (overexposed images)  
# lapply(tail_results_list, function(x) open_tail_urls(x, open_dark = FALSE, open_bright = TRUE))

# # Open all tail URLs
# lapply(tail_results_list, function(x) open_tail_urls(x, open_dark = TRUE, open_bright = TRUE))

# === SUMMARY STATISTICS ===
cat("\n", rep("=", 80), "\n", sep="")
cat("SUMMARY OF TAIL ANALYSIS\n")
cat(rep("=", 80), "\n", sep="")

summary_df <- data.frame(
  species = sapply(tail_results_list, function(x) x$species),
  cell = sapply(tail_results_list, function(x) x$cell), 
  dark_threshold = sapply(tail_results_list, function(x) x$dark_tail$threshold),
  dark_count = sapply(tail_results_list, function(x) x$dark_tail$n_images),
  bright_threshold = sapply(tail_results_list, function(x) x$bright_tail$threshold),
  bright_count = sapply(tail_results_list, function(x) x$bright_tail$n_images),
  stringsAsFactors = FALSE
)

print(summary_df)

# Return the tail results for further analysis
invisible(tail_results_list)
```

Plot illumination bias (skew) over the contiguous USA

```{r}

# Test if illumination bias is geographically structured
geo_data <- d %>% 
  distinct(cell, .keep_all = TRUE) %>%
  select(cell, cell_lat, cell_lon, illumination_index, cell_skew_weighted, lightness_skew,
         cell_cv_weighted, illumination_variance_index)

# Correlation with latitude and longitude
geo_cor <- cor(geo_data[, c("cell_lat", "cell_lon", "illumination_index", 
                             "cell_skew_weighted", "cell_cv_weighted","lightness_skew")],
               use = "complete.obs")
print(round(geo_cor, 3))

# Visualize geographic pattern
p1 <- ggplot(geo_data, aes(x = cell_lon, y = cell_lat, color = lightness_skew)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                        name = "Illumination\nIndex") +
  labs(title = "Geographic Distribution of Illumination Bias",
       subtitle = "Red = likely underexposed, Blue = likely overexposed",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
print(p1)

# Test for spatial autocorrelation
library(ape)
coords <- as.matrix(geo_data[, c("cell_lon", "cell_lat")])
dists <- as.matrix(dist(coords))
dists.inv <- 1/dists
diag(dists.inv) <- 0

# Moran's I test
Moran.I(geo_data$illumination_index, dists.inv, na.rm = TRUE)

# Pull out one row per cell with illumination data
geo_data <- d %>%
  distinct(cell, .keep_all = TRUE) %>%
  select(cell, cell_lat, cell_lon, lightness_skew)

world <- map_data("world")
ggplot() +
  geom_map(
    data = world, map = world,
    aes(map_id = region),
    fill = "grey90", color = "white", linewidth = 0.2
  ) +
  geom_point(
    data = geo_data,
    aes(x = cell_lon, y = cell_lat, color = lightness_skew),
    size = 3, alpha = 0.8
  ) +
  scale_color_gradient2(
    low = "black", mid = "grey", high = "white", 
    midpoint = 0,
    name = "Lightness\nSkew"
  ) +
  coord_quickmap(xlim = range(geo_data$cell_lon),
                 ylim = range(geo_data$cell_lat)) +
  labs(title = "Geographic Distribution of Illumination Bias",
       subtitle = "Red = underexposed bias, Blue = overexposed bias") +
  theme_minimal()

geo_data <- d %>% 
  distinct(cell, .keep_all = TRUE) %>%
  select(cell, cell_lat, cell_lon, cell_skew_weighted, illumination_index, 
         cell_cv_weighted)

# Then plot with cell_skew_weighted instead
p1 <- ggplot(geo_data, aes(x = cell_lon, y = cell_lat, color = cell_skew_weighted))
```
