---
title: "publication_figures"
---

Data and libraries

```{r}

library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(tidyr)
library(colorEvoHelpers)
library(patchwork)
library(grid)

d <- readRDS("_targets/objects/d_sp_cell")
```

Figure 2. Distribution of data on map with tiny hists

```{r}

plotTinyHist <- function(df){
  # 1) summarize with .groups="drop"
  cells <- df %>%
    group_by(cell_lat, cell_lon) %>%
    summarise(
      cell_id = Mode(cell),
      n       = n(),
      nspecies= n_distinct(species),
      nApril  = sum(onset >=  91 & onset <= 120, na.rm = TRUE),
      nMay    = sum(onset >= 121 & onset <= 151, na.rm = TRUE),
      nJune   = sum(onset >= 152 & onset <= 181, na.rm = TRUE),
      nJuly   = sum(onset >= 182 & onset <= 212, na.rm = TRUE),
      nAugust = sum(onset >= 213 & onset <= 243, na.rm = TRUE),
      .groups = "drop"
    )

  # compute scales
  lon_rng   <- range(cells$cell_lon)
  lat_rng   <- range(cells$cell_lat)
  highest_n <- max(cells$n)

  # pivot
  month_colors <- c(
    nApril  = "#377eb8", nMay    = "#66c2a5",
    nJune   = "#c9c952", nJuly   = "#ff7f00",
    nAugust = "#e41a1c"
  )
  cells_long <- cells %>%
    pivot_longer(names_to = "month", values_to = "count",
                 cols = names(month_colors)) %>%
    mutate(
      month_idx = match(month, names(month_colors)),
      total_w   = (lon_rng[2] - lon_rng[1]) / 30,
      total_h   = (lat_rng[2] - lat_rng[1]) / 5,
      bar_w     = total_w / length(month_colors),
      xmin      = cell_lon - total_w/2 + (month_idx-1)*bar_w,
      xmax      = xmin + bar_w,
      ymin      = cell_lat,
      ymax      = cell_lat + (count/highest_n)*total_h
    )

  # base map with linewidth
  world <- map_data('world')
  p <- ggplot(world, aes(long, lat)) +
    geom_map(
      map      = world,
      aes(map_id = region),
      fill     = "grey95",
      color    = "white",
      linewidth= 0.2
    ) +
    coord_quickmap(xlim = lon_rng, ylim = lat_rng) +
    theme_minimal()

  # tiny histograms, turning off inheritance of long/lat aes
  p + geom_rect(
        data         = cells_long,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = month),
        inherit.aes  = FALSE
      ) +
      scale_fill_manual(
    values = month_colors,
    breaks = names(month_colors),               # <- enforce the order
    labels = c("Apr","May","Jun","Jul","Aug")   # <- these labels now align
  ) +
      guides(fill = guide_legend(title = "Onset Month")) +
      labs(
        x         = NULL, y = NULL
      ) +
      theme(
        panel.grid  = element_blank(),
        axis.text   = element_blank(),
        axis.ticks  = element_blank()
      )
}

plotTinyHist(d)

```

Fig. 2. Distribution of data on map with tiny hists - finding values for a cell

```{r}

# pull out one row per cell
cells <- d %>%
  distinct(cell, cell_lat, cell_lon)

world <- map_data("world")
ggplot() +
  geom_map(
    data = world, map = world,
    aes(map_id = region),
    fill = "grey90", color = "white", linewidth = 0.2
  ) +
  geom_text(
    data = cells,
    aes(x = cell_lon, y = cell_lat, label = cell),
    size = 3, color = "black"
  ) +
  coord_quickmap(xlim = range(cells$cell_lon),
                 ylim = range(cells$cell_lat)) +
  theme_minimal()
```

Figure 3. Latitude models

```{r}

library(lme4)
library(ggeffects)
library(splines)
library(sjPlot)
library(ggplot2)

l_m4 <- readRDS("models/l_m4.rds")
l_m2 <- readRDS("models/l_m2.rds")

# l_m4 is the best model by AIC, but m2 is still interesting
p <- plot_model(
  l_m4,
  type      = "pred",
  terms     = c("cell_lat_sc", "onset"),
  show.data = TRUE
)
# m2 was the best model by AIC
p <- plot_model(
  l_m2,
  type      = "pred",
  terms     = c("cell_lat_sc", "onset_sc [-2, 0, 2]"),
  show.data = TRUE
)

# recolor layer 2 (the data points) to black:
p$layers[[1]]$aes_params$colour <- "black"
p$layers[[1]]$aes_params$alpha <- 0.05
p <- p +
  labs(
    title  = NULL,            # remove title
    x      = "Scaled latitude",      # new x label
    y      = "Mean dark body coverage",  # new y label
    colour = "Onset Day-of-Year"   # new legend title
  ) +
  theme(
    plot.title = element_blank()  # make sure title is gone
  )

p
```

Figure 3. Latitude models - getting actual dates to put on legend

```{r}

# 1) grab the centering & scaling used for onset_corrected
sc_info      <- attributes(scale(d_sp_cell$onset_corrected))
onset_center <- sc_info$`scaled:center`
onset_scale  <- sc_info$`scaled:scale`

# 2) your three scaled values:
onset_sc_vals <- c(-2, 0, 2)

# 3) recover the day‐of‐year (DOY) for each
doy_vals <- round(onset_sc_vals * onset_scale + onset_center)

# 4) convert DOY into actual dates (using a non‐leap reference year—here 2025)
dates <- as.Date(doy_vals - 1, origin = "2025-01-01")

# 5) format as month & day
month_day <- format(dates, "%B %d")

# 6) put it all in a little table
data.frame(
  onset_sc   = onset_sc_vals,
  DOY        = doy_vals,
  date       = dates,
  month_day  = month_day
)
```

Figure 4. Climate variable models

```{r}

data <- read.csv("D:/bcc/chloros/records.csv")
length(unique(data$species))

# back transform date

m1 <- readRDS("models/m1.rds")
m2 <- readRDS("models/m2.rds")
m3 <- readRDS("models/m3.rds")

# generate each panel
p1 <- plot_model(m1,type="pred",terms=c("tmean_sc","onset_sc")) + ylim(0.2, 0.46)
p2 <- plot_model(m1,type="pred",terms=c("srad_sc")) + ylim(0.2, 0.46) + labs(color = "Onset") + xlim(-2,2)

p3 <- plot_model(m2,type="pred",terms=c("tmean_sc","onset_sc")) + ylim(0.2, 0.46)
p3
p4 <- plot_model(m2,type="pred",terms=c("vp_sc")) + ylim(0.2, 0.46) + labs(color = "Onset") + xlim(-2,2)

p5 <- plot_model(m3,type="pred",terms=c("vp_sc","onset_sc")) + ylim(0.2, 0.46)
p6 <- plot_model(m3,type="pred",terms=c("srad_sc")) + ylim(0.2, 0.46) + labs(color = "Onset") + xlim(-2,2)

# Add a blank spacer between rows
spacer <- plot_spacer() + plot_layout(height = 0.05)

# Combine with spacer in the middle
combined <- (p1 | p2) / spacer / (p3 | p4) / spacer / (p5 | p6) +
  plot_layout(guides = "collect",
              heights = c(1, 0.05, 1, 0.05, 1))

combined <- combined &
  theme(plot.title = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())
combined
```

Attempt to plot latitude effect on map, but doesn't look great

```{r}

plotColorOnMap <- function(onset_min=0,onset_max=360,lat_min=35,my_species=NA,data=d){
  
  library(ggplot2)
  library(rnaturalearth)
  library(rnaturalearthdata)
  library(dplyr)
  library(sf)
  
  d <- data
  d <- dplyr::filter(d, onset > onset_min)
  d <- dplyr::filter(d, onset < onset_max)
  d <- dplyr::filter(d, cell_lat > lat_min)
  if(!is.na(my_species)){
    d <- dplyr::filter(d, species == my_species)
  }
  # Base world map
  world <- map_data("world")
  
  # Ensure df has cell_lat, cell_lon, black_20
  ggplot() +
  geom_map(
    data = world,
    map = world,
    aes(long, lat, map_id = region),
    fill = "grey95",
    color = "white",
    linewidth = 0.2
  ) +
  geom_point(
    data = d,
    aes(x = cell_lon, y = cell_lat, fill = black_20),
    shape = 21,
    color = "black",
    size = 5
  ) +
  scale_fill_gradient(
    low = "white", high = "black",
    name = "Darkness (black_20)"
  ) +
  coord_quickmap(
  xlim = range(d$cell_lon, na.rm = TRUE),
  ylim = range(d$cell_lat, na.rm = TRUE)
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "right"
  ) +
  labs(x = NULL, y = NULL)
}

hist(d$onset)
plotColorOnMap(onset_min=-3,onset_max=-1,lat_min=35,data=d)
plotColorOnMap(onset_min=-3,onset_max=0,lat_min=35,my_species="Basiaeschna janata",data=d)
plotColorOnMap(onset_min=0,onset_max=3,lat_min=35,my_species="Basiaeschna janata",data=d)

plotColorOnMap(onset_min=0,onset_max=1,lat_min=35,data=d)
plotColorOnMap(onset_min=1,onset_max=3,data=d)

table(d$species)
```

```{r}

# fit a linear model of how observation count explains onset
# more observers could mean earlier onset 

```

```{r}

library(sjPlot)
library(ggplot2)
library(patchwork)

# --- same setup you already have for labels/colors ---
mu_onset <- if (!is.null(attr(d$onset_sc, "scaled:center"))) attr(d$onset_sc, "scaled:center") else mean(d$onset, na.rm = TRUE)
sd_onset <- if (!is.null(attr(d$onset_sc, "scaled:scale")))  attr(d$onset_sc, "scaled:scale")  else sd(d$onset,  na.rm = TRUE)

onset_label_fun_vec <- function(keys_chr) {
  sc  <- as.numeric(keys_chr)
  doy <- sc * sd_onset + mu_onset
  format(as.Date(doy - 1, origin = "2001-01-01"), "%b %d")
}
onset_lvls <- as.numeric(quantile(d$onset_sc, c(.1, .5, .9), na.rm = TRUE))
onset_keys <- sprintf("%.2f", onset_lvls)
onset_term <- paste0("onset_sc[", paste(onset_keys, collapse=","), "]")
onset_cols <- setNames(c("#3B528B", "#80B29E","#D73027"), onset_keys)  # early=blue, mid=gray, late=red
onset_labs <- onset_label_fun_vec(onset_keys)

# --- apply both color AND fill scales to grouped plots (p1, p3, p5) ---
p1 <- plot_model(m1, type = "pred",
                 terms = c("tmean_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")  # keep legend once

p2 <- plot_model(m1, type = "pred",
                 terms = "srad_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)

p3 <- plot_model(m2, type = "pred",
                 terms = c("tmean_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")

p4 <- plot_model(m2, type = "pred",
                 terms = "vp_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)

p5 <- plot_model(m3, type = "pred",
                 terms = c("vp_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")

p6 <- plot_model(m3, type = "pred",
                 terms = "srad_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)

p2 <- p2 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())

p4 <- p4 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())

p6 <- p6 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())

# X-axis titles
p1 <- p1 + labs(x = "Temperature")
p2 <- p2 + labs(x = "Solar Radiation")
p3 <- p3 + labs(x = "Temperature")
p4 <- p4 + labs(x = "Vapor Pressure")
p5 <- p5 + labs(x = "Vapor Pressure")
p6 <- p6 + labs(x = "Solar Radiation")

# ---- layout ----
spacer <- plot_spacer() + plot_layout(height = 0.05)

combined <- (p1 | p2) / spacer / (p3 | p4) / spacer / (p5 | p6) +
  plot_layout(guides = "collect",
              heights = c(1, 0.05, 1, 0.05, 1))

combined <- combined & theme(
  plot.title   = element_blank(),
  axis.title.y = element_blank()  # keep or remove as you like
  # REMOVE this line so x titles show:
  # axis.title.x = element_blank()
)
combined
```

Find extremes in intraspecific variation in melanism

```{r}

library(dplyr)
library(ggplot2)

# Summarize intraspecific variation in melanism per species
sp_var <- d %>%
  group_by(species) %>%
  summarise(
    n_obs       = n(),
    n_cells     = n_distinct(cell),
    mean_black  = mean(black_20, na.rm = TRUE),
    sd_black    = sd(black_20, na.rm = TRUE),
    cv_black    = sd_black / mean_black,
    range_black = diff(range(black_20, na.rm = TRUE)),
    q10         = quantile(black_20, 0.10, na.rm = TRUE),
    q90         = quantile(black_20, 0.90, na.rm = TRUE),
    iqr_black   = q90 - q10,
    lat_range   = diff(range(cell_lat, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  # require reasonable sample size & geographic spread so extremes aren't noise
  filter(n_obs >= 10, n_cells >= 5)

# Top 10 most variable species (by SD, robust to outliers via IQR too)
top_sd  <- sp_var %>% arrange(desc(sd_black))  %>% slice_head(n = 10)
top_iqr <- sp_var %>% arrange(desc(iqr_black)) %>% slice_head(n = 10)
top_cv  <- sp_var %>% arrange(desc(cv_black))  %>% slice_head(n = 10)

# Least variable species for comparison
bot_sd  <- sp_var %>% arrange(sd_black) %>% slice_head(n = 10)

cat("=== Most variable (SD of black_20) ===\n")
print(top_sd  %>% select(species, n_obs, n_cells, mean_black, sd_black, iqr_black, lat_range))

cat("\n=== Most variable (10–90% IQR) ===\n")
print(top_iqr %>% select(species, n_obs, n_cells, mean_black, sd_black, iqr_black, lat_range))

cat("\n=== Most variable (coefficient of variation) ===\n")
print(top_cv  %>% select(species, n_obs, n_cells, mean_black, sd_black, cv_black, lat_range))

cat("\n=== Least variable (SD) ===\n")
print(bot_sd  %>% select(species, n_obs, n_cells, mean_black, sd_black, iqr_black, lat_range))

# Visualize: distribution of black_20 for the top 10 most variable species
top_species <- top_sd$species

d_top <- d %>%
  filter(species %in% top_species) %>%
  mutate(species = factor(species, levels = top_species))  # order by SD

ggplot(d_top, aes(x = reorder(species, black_20, FUN = sd),
                  y = black_20, fill = species)) +
  geom_violin(alpha = 0.6, color = NA, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.8) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Dark body coverage (black_20)",
    title = "Species with greatest intraspecific variation in melanism"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(face = "italic"))

```

```{r}

library(dplyr)
library(purrr)
library(tidyr)

# Unnest photos + lightness values so each row = one photo
d_photos <- d %>%
  select(cell, cell_lat, cell_lon, species, urls, lightness_values, ids) %>%
  mutate(row_id = row_number()) %>%
  unnest(cols = c(urls, lightness_values, ids))

cell_representatives <- d_photos %>%
  #filter(species == "Sympetrum semicinctum") %>%
  #filter(species == "Cordulegaster maculata") %>%
  #filter(species == "Progomphus obscurus") %>%
  #filter(species == "Epitheca cynosura") %>%
  #filter(species == "Pantala hymenaea") %>%
  #filter(species == "Pachydiplax longipennis") %>%
  #filter(species == "Anax junius") %>%
  #filter(species == "Tramea lacerata") %>%
  #filter(species == "Celithemis eponina") %>%
  filter(species == "Boyeria vinosa") %>%

  group_by(cell) %>%
  mutate(
    cell_mean = mean(lightness_values, na.rm = TRUE),
    dist_mean = abs(lightness_values - cell_mean)
  ) %>%
  slice_min(dist_mean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(cell_lat, cell_lon) %>%
  select(cell, cell_lat, cell_lon, species, cell_mean, lightness_values, urls, ids)

print(cell_representatives, n = Inf)

# Top-k closest photos per cell (optional)
top_k_per_cell <- d_photos %>%
  group_by(cell) %>%
  mutate(
    cell_mean = mean(lightness_values, na.rm = TRUE),
    dist_mean = abs(lightness_values - cell_mean)
  ) %>%
  slice_min(dist_mean, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(cell, dist_mean) %>%
  select(cell, cell_lat, cell_lon, species, cell_mean, lightness_values, urls)

# Quick printout for browsing
cell_representatives %>%
  mutate(out = sprintf("cell %s | mean=%.2f | this=%.2f | %s | id=%s | %s",
                       cell, cell_mean, lightness_values, species, ids, urls)) %>%
  pull(out) %>%
  cat(sep = "\n")
```

Fig 5 chunk with marginal

```{r}

library(sjPlot)
library(ggplot2)
library(patchwork)
library(emmeans)

# --- same setup you already have for labels/colors ---
mu_onset <- if (!is.null(attr(d$onset_sc, "scaled:center"))) attr(d$onset_sc, "scaled:center") else mean(d$onset, na.rm = TRUE)
sd_onset <- if (!is.null(attr(d$onset_sc, "scaled:scale")))  attr(d$onset_sc, "scaled:scale")  else sd(d$onset,  na.rm = TRUE)
onset_label_fun_vec <- function(keys_chr) {
  sc  <- as.numeric(keys_chr)
  doy <- sc * sd_onset + mu_onset
  format(as.Date(doy - 1, origin = "2001-01-01"), "%b %d")
}
onset_lvls <- as.numeric(quantile(d$onset_sc, c(.1, .5, .9), na.rm = TRUE))
onset_keys <- sprintf("%.2f", onset_lvls)
onset_term <- paste0("onset_sc[", paste(onset_keys, collapse=","), "]")
onset_cols <- setNames(c("#3B528B", "#80B29E","#D73027"), onset_keys)
onset_labs <- onset_label_fun_vec(onset_keys)

# --- estimated marginal trends of climate at each onset level ---
# M1: temperature slope at each onset level
cat("=== M1: Temperature trends by onset ===\n")
m1_tmean_trends <- emtrends(m1, ~onset_sc, var = "tmean_sc",
                            at = list(onset_sc = onset_lvls))
print(summary(m1_tmean_trends))

# M1: solar radiation slope (no interaction with onset, but for completeness)
cat("\n=== M1: Solar radiation trends by onset ===\n")
m1_srad_trends <- emtrends(m1, ~onset_sc, var = "srad_sc",
                           at = list(onset_sc = onset_lvls))
print(summary(m1_srad_trends))

# M2: temperature slope at each onset level
cat("\n=== M2: Temperature trends by onset ===\n")
m2_tmean_trends <- emtrends(m2, ~onset_sc, var = "tmean_sc",
                            at = list(onset_sc = onset_lvls))
print(summary(m2_tmean_trends))

# M2: humidity slope at each onset level
cat("\n=== M2: Humidity trends by onset ===\n")
m2_vp_trends <- emtrends(m2, ~onset_sc, var = "vp_sc",
                         at = list(onset_sc = onset_lvls))
print(summary(m2_vp_trends))

# M3: solar radiation slope at each onset level
cat("\n=== M3: Solar radiation trends by onset ===\n")
m3_srad_trends <- emtrends(m3, ~onset_sc, var = "srad_sc",
                           at = list(onset_sc = onset_lvls))
print(summary(m3_srad_trends))

# M3: humidity slope at each onset level
cat("\n=== M3: Humidity trends by onset ===\n")
m3_vp_trends <- emtrends(m3, ~onset_sc, var = "vp_sc",
                         at = list(onset_sc = onset_lvls))
print(summary(m3_vp_trends))

# --- figure 5 plots ---
p1 <- plot_model(m1, type = "pred",
                 terms = c("tmean_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")
p2 <- plot_model(m1, type = "pred",
                 terms = "srad_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)
p3 <- plot_model(m2, type = "pred",
                 terms = c("tmean_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")
p4 <- plot_model(m2, type = "pred",
                 terms = "vp_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)
p5 <- plot_model(m3, type = "pred",
                 terms = c("vp_sc[-2:2 by=0.5]", onset_term)) +
      ylim(0.2, 0.46) +
      scale_color_manual(name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      scale_fill_manual (name = "Onset Date", breaks = onset_keys, labels = onset_labs, values = onset_cols) +
      guides(fill = "none")
p6 <- plot_model(m3, type = "pred",
                 terms = "srad_sc[-2:2 by=0.5]") +
      ylim(0.2, 0.46)
p2 <- p2 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())
p4 <- p4 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())
p6 <- p6 + theme(axis.text.y = element_blank(),
                 axis.ticks.y = element_blank(),
                 axis.title.y = element_blank())
p1 <- p1 + labs(x = "Temperature")
p2 <- p2 + labs(x = "Solar Radiation")
p3 <- p3 + labs(x = "Temperature")
p4 <- p4 + labs(x = "Vapor Pressure")
p5 <- p5 + labs(x = "Vapor Pressure")
p6 <- p6 + labs(x = "Solar Radiation")

spacer <- plot_spacer() + plot_layout(height = 0.05)
combined <- (p1 | p2) / spacer / (p3 | p4) / spacer / (p5 | p6) +
  plot_layout(guides = "collect",
              heights = c(1, 0.05, 1, 0.05, 1))
combined <- combined & theme(
  plot.title   = element_blank(),
  axis.title.y = element_blank()
)
combined

```
