---
title: "Supplemental Sensitivity"
---

Analysis 1: Skew removed

Uses threshold darkness metric values derived from distributions with skew removed

To run: run the first code chunk below, then the second chunk from model_selection.qmd, then the second chunk below to save the models.

```{r}

library(colorEvoHelpers)
library(dplyr)
library(tidyr)
library(phyr)
library(dplyr)
library(car)
library(performance)
library(MuMIn)
library(buildmer)
library(glmmTMB)
library(splines)
library(sjPlot)
library(ggeffects)
library(brms)
library(sjPlot)

d <- readRDS("_targets/objects/d_sp_cell")

d$black_20 <- d$black_20_sens 
threshold_str <- "black_20 ~"
```

```{r}

saveRDS(m1, "models/m1_skewrm.rds")
saveRDS(m2, "models/m2_skewrm.rds")
saveRDS(m3, "models/m3_skewrm.rds")
```

Analysis 2: Black 30 threshold instead of black 20

Uses a more permissive threshold

To run: run the first code chunk below, then the second chunk from model_selection.qmd, then the second chunk below to save the models.

```{r}

library(colorEvoHelpers)
library(dplyr)
library(tidyr)
library(phyr)
library(dplyr)
library(car)
library(performance)
library(MuMIn)
library(buildmer)
library(glmmTMB)
library(splines)
library(sjPlot)
library(ggeffects)
library(brms)
library(sjPlot)

d <- readRDS("_targets/objects/d_sp_cell")
threshold_str <- "black_30 ~"
```

```{r}

saveRDS(m1, "models/m1_black30.rds")
saveRDS(m2, "models/m2_black30.rds")
saveRDS(m3, "models/m3_black30.rds")
```

Analysis 2: Strong phylogenetic correction

To run: run ALL the chunks from model_selection.qmd, then run the below chunk

```{r}

# Wrapper to refit a glmmTMB model with phylogeny via phyr
refitWithPhylo <- function(glmmTMB_model, data, tree) {
  # Extract formula and modify for phyr syntax
  f <- formula(glmmTMB_model)
  f_char <- as.character(f)
  
  # Replace (1 | species) with (1 | species__)
  f_char[3] <- gsub("\\(1 \\| species\\)", "(1 | species__)", f_char[3])
  new_formula <- as.formula(paste(f_char[2], "~", f_char[3]))
  
  phyr::pglmm(
    new_formula,
    data = data,
    cov_ranef = list(species = tree),
    REML = FALSE
  )
}

tree_trimmed <- targets::tar_read(tree_trimmed)

# Compare
m1 <- refitWithPhylo(m1, d, tree_trimmed)
m2 <- refitWithPhylo(m2, d, tree_trimmed)
m3 <- refitWithPhylo(m3, d, tree_trimmed)

saveRDS(m1, "models/m1_phy.rds")
saveRDS(m2, "models/m2_phy.rds")
saveRDS(m3, "models/m3_phy.rds")
```

Read all and summarize to assess similarity to base model

```{r}

# Load all sensitivity analysis models
m1_skewrm <- readRDS("models/m1_skewrm.rds")
m2_skewrm <- readRDS("models/m2_skewrm.rds")
m3_skewrm <- readRDS("models/m3_skewrm.rds")

m1_black30 <- readRDS("models/m1_black30.rds")
m2_black30 <- readRDS("models/m2_black30.rds")
m3_black30 <- readRDS("models/m3_black30.rds")

m1_phy <- readRDS("models/m1_phy.rds")
m2_phy <- readRDS("models/m2_phy.rds")
m3_phy <- readRDS("models/m3_phy.rds")

# Load base models for comparison
m1 <- readRDS("models/m1.rds")
m2 <- readRDS("models/m2.rds")
m3 <- readRDS("models/m3.rds")

# Summarize all models
cat("========== BASE MODELS ==========\n\n")
cat("--- Model 1 (Base) ---\n")
summary(m1)

cat("\n--- Model 2 (Base) ---\n")
summary(m2)

cat("\n--- Model 3 (Base) ---\n")
summary(m3)

cat("\n\n========== SKEW REMOVED ==========\n\n")
cat("--- Model 1 (Skew Removed) ---\n")
summary(m1_skewrm)

cat("\n--- Model 2 (Skew Removed) ---\n")
summary(m2_skewrm)

cat("\n--- Model 3 (Skew Removed) ---\n")
summary(m3_skewrm)

cat("\n\n========== BLACK 30 THRESHOLD ==========\n\n")
cat("--- Model 1 (Black 30) ---\n")
summary(m1_black30)

cat("\n--- Model 2 (Black 30) ---\n")
summary(m2_black30)

cat("\n--- Model 3 (Black 30) ---\n")
summary(m3_black30)

cat("\n\n========== PHYLOGENETIC CORRECTION ==========\n\n")
cat("--- Model 1 (Phylogenetic) ---\n")
summary(m1_phy)

cat("\n--- Model 2 (Phylogenetic) ---\n")
summary(m2_phy)

cat("\n--- Model 3 (Phylogenetic) ---\n")
summary(m3_phy)
```
