---
title: "new_models"
---

Data are species 250 km cell combos (not seasonal or monthly)

Get phenesse seasonal weib onset and offset

Get cell average temperature at month of onset and offset - temperature context of onset time point

Model looks like: black \~ onset_of_species_in_cell \* latitude_of_cell + temperature_at_onset_of_species_in_cell + srad_at_onset_of_species_in_cell + (1\|species) + (1\|cell)

Could add species level slopes for temperature

Could add percher flier

Could add body size

This leads fixed effects to take on more explanatory power, less in random effect

10 records per cell is minimum for phenesse

First create data

```{r}

library(phenesse)
library(dplyr)
library(tidyr)
library(colorEvoHelpers)

d_gridded <- readRDS("_targets/objects/d_gridded")
d_gridded <- d_gridded %>% dplyr::select(-geometry)
d_gridded <- d_gridded %>%
  distinct(img_id, .keep_all = TRUE)
d_gridded <- d_gridded %>% filter(yday>0)

d_sp_cell <- d_gridded %>%
  group_by(species, cell) %>%
  filter(n() >= 10) %>%
  group_modify(~ {
    # cap at 50 
    if (nrow(.x) > 50) {
      .x <- slice_sample(.x, n = 50)
    }

    # compute onset/offset
    tibble(
      onset = phenesse::quantile_ci(.x$yday,0.1),
      #onset = quantile(.x$yday,0.1),

      cell_lat = Mode(.x$cell_lat),
      cell_lon = Mode(.x$cell_lon),
      year = Mode(.x$year),
      black_20 = mean(.x$hls_ch1_below0.2),
      black_30 = mean(.x$hls_ch1_below0.3),
      mean_lightness = mean(.x$mean_cielab_lightness),
      wing_length = mean(.x$sp_winglength),
      flight_type = Mode(.x$flight_type),
      n = nrow(.x)
    )
  }) %>%
  ungroup()

# correct onset based on n obs
t <- lm(onset ~ n, data = d_sp_cell)
summary(t)
bias <- predict(t)
d_sp_cell$onset_corrected <- d_sp_cell$onset - bias + mean(d_sp_cell$onset)
hist(d_sp_cell$onset, main = "Original onset", col = "gray")
hist(onset_corrected, main = "Corrected onset", col = "steelblue")

saveRDS(d_sp_cell,"d_sp_cell.rds")

# subtract corrected from uncorrected 
# for species cell combos with less sampling should be pushed backward in time when corrected
```

Download prism data if you need

```{r}

library(prism)

prism_set_dl_dir("D:/prism")
# 2a) Download all DAILY tmean rasters from Jan 1 2017 through Dec 31 2023
get_prism_monthlys(
  type     = "vpdmax",
  years = 2021:2024
)
get_prism_monthlys(
  type     = "vpdmin",
  years = 2021:2024
)
get_prism_monthlys(
  type     = "tmean",
  years = 2021:2024
)
```

Bind annual humidity data

```{r}

library(colorEvoHelpers)
d_sp_cell$season <- d_sp_cell$year
d_sp_cell <- bindPrismData(d_sp_cell,variable="vpdmin",prism_dir="D:/prism",avg_of_years=2021:2023,seasonal=FALSE,new_var_colname="vpdmin")

d_sp_cell <- bindPrismData(d_sp_cell,variable="vpdmax",prism_dir="D:/prism",avg_of_years=2021:2023,seasonal=FALSE,new_var_colname="vpdmax")

d_sp_cell <- bindPrismData(d_sp_cell,variable="tmean",prism_dir="D:/prism",avg_of_years=2021:2023,seasonal=FALSE,new_var_colname="tmean")

d_sp_cell <- bindBioclimData(d_sp_cell,bioclim_var_num = 1, new_var_colname="srad", res=10,path="D:/bioclim_data")
saveRDS(d_sp_cell,"d_sp_cell2.rds")

#unique(d_sp_cell$species)
# could filter species with less than 5 cells


```

```{r}

bindWorldClimData <- function(df, var = "bio", bioclim_var_num = 1, new_var_colname = "tmax", 
                              res = 10, path = 'D:/bioclim_data', month = NULL, cellsize_km = NULL) {
  
  library(geodata)
  library(terra)
  library(dplyr)
  
  # Download WorldClim data
  clim_data <- worldclim_global(var = var, res = res, path = path)
  
  # Choose the appropriate layer
  if (var == "bio") {
    selected_layer <- clim_data[[bioclim_var_num]]
  } else if (var %in% c("srad", "vapr", "tavg", "tmin", "tmax", "wind")) {
    if (is.null(month)) {
      selected_layer <- mean(clim_data)
    } else {
      selected_layer <- clim_data[[month]]
    }
  } else {
    stop("Unsupported variable type. Use 'bio', 'srad', 'vapr', 'tavg', 'tmin', 'tmax', or 'wind'.")
  }
  
  # Convert df to SpatVector
  points <- vect(df, geom = c("longitude", "latitude"), crs = crs(selected_layer))
  
  # If a buffer size is specified, buffer the points and take mean
  if (!is.null(cellsize_km)) {
    radius_m <- (cellsize_km * 1000) / 2  # convert to radius in meters
    buffered_points <- buffer(points, width = radius_m)
    extracted_data <- extract(selected_layer, buffered_points, fun = mean, na.rm = TRUE)
  } else {
    # Just extract the value at the point
    extracted_data <- extract(selected_layer, points)
  }
  
  # Combine with original dataframe
  result_df <- cbind(df, extracted_data[, -1])
  
  # Rename new column
  last_col_name <- names(result_df)[ncol(result_df)]
  result_df <- result_df %>% rename(!!new_var_colname := all_of(last_col_name))
  
  return(result_df)
}

d_sp_cell$latitude <- d_sp_cell$cell_lat
d_sp_cell$longitude <- d_sp_cell$cell_lon
d_sp_cell <- bindWorldClimData(d_sp_cell, var = "tavg", new_var_colname = "tmean")
d_sp_cell <- bindWorldClimData(d_sp_cell, var = "srad", new_var_colname = "srad")
d_sp_cell <- bindWorldClimData(d_sp_cell, var = "vapr", new_var_colname = "vp")

```

Option - bind on monthly temperature and srad data

```{r}

#regarding vpd not seeing strong effects so far, given that vpd and lat are correlated I think that lat spline is sucking up a lot of the variation

library(dplyr)
library(colorEvoHelpers)

## Helper: build cell → polygon lookup
makeCellPolygons <- function(df, cell_width_km = 250) {
  df %>%
    distinct(cell, cell_lat, cell_lon) %>%
    rowwise() %>%
      mutate(
        cell_poly = list(
          cellPolyFromLatLonWidth(cell_lat, cell_lon, cell_width_km)
        )
      ) %>%
    ungroup()
}

## Helper: fetch PRISM data for one cell & one month or year
fetch_prism_for_cell <- function(
  type,
  temp_period = c("monthly","annual","daily"),
  years       = NULL,
  mon         = NULL,
  minDate     = NULL,
  maxDate     = NULL,
  cell_poly
) {
  temp_period <- match.arg(temp_period)
  args <- list(type = type, temp_period = temp_period)
  if (temp_period == "monthly") {
    args$years <- years; args$mon <- mon
  } else if (temp_period == "annual") {
    args$years <- years
  } else {
    args$minDate <- as.character(minDate)
    args$maxDate <- as.character(maxDate)
  }

  pd    <- do.call(prism_archive_subset, args)
  files <- pd_to_file(pd)
  ras   <- stack(lapply(files, raster))
  avg   <- calc(ras, fun = mean)
  cropped <- crop(avg, extent(cell_poly))
  masked  <- mask(cropped, cell_poly)
  cellStats(masked, stat = "mean")
}

## Main: bind PRISM monthly values per-row using a 'month' column
bindPrismByMonth <- function(
  df,
  prism_dir,
  month_col,
  years           = 2015:2023,
  cell_width_km   = 250,
  type            = "tmean",
  new_var_colname = "prism_tmean"
) {
  library(dplyr)
  library(prism)
  library(raster)
  library(geosphere)
  library(purrr)

  # 1) set PRISM download dir
  prism_set_dl_dir(prism_dir)

  # 2) build cell → polygon lookup
  polys <- makeCellPolygons(df, cell_width_km)

  # 3) derive unique tasks: each cell-year-month
  tasks <- df %>%
    filter(year %in% years) %>%
    mutate(
      mon = .data[[month_col]]
    ) %>%
    distinct(cell, year, mon) %>%
    left_join(polys, by = "cell")

  # 4) fetch PRISM mean for each task
  results <- tasks %>%
    mutate(
      value = pmap_dbl(
        list(year, mon, cell_poly),
        function(yr, m, poly) {
          fetch_prism_for_cell(
            type        = type,
            temp_period = "monthly",
            years       = yr,
            mon         = m,
            cell_poly   = poly
          )
        }
      )
    ) %>%
    dplyr::select(cell, year, mon, value)

  # 5) join back to original df
  df %>%
    left_join(
      results %>% rename(!!month_col := mon),
      by = c("cell", "year", month_col)
    ) %>%
    rename(!!new_var_colname := value)
}

# version that works with avg of years
bindPrismByMonth <- function(
  df,
  prism_dir,
  month_col,
  years           = 2021:2023,
  cell_width_km   = 250,
  type            = "tmean",
  new_var_colname = "prism_tmean"
) {
  library(dplyr)
  library(prism)
  library(raster)
  library(geosphere)
  library(purrr)

  # 1) set PRISM download dir
  prism_set_dl_dir(prism_dir)

  # 2) build cell → polygon lookup once
  polys <- makeCellPolygons(df, cell_width_km)

  # 3) derive unique tasks: each cell × month (ignore year)
  tasks <- df %>%
    mutate(
      mon = .data[[month_col]]
    ) %>%
    distinct(cell, mon) %>%
    left_join(polys, by = "cell")

  # 4) fetch PRISM mean across ALL years for each (cell,mon)
  results <- tasks %>%
    mutate(
      value = map2_dbl(
        mon, cell_poly,
        ~ fetch_prism_for_cell(
            type        = type,
            temp_period = "monthly",
            years       = years,   # pass full vector here!
            mon         = .x,
            cell_poly   = .y
          )
      )
    ) %>%
    dplyr::select(cell, mon, value)

  # 5) join back to original df by cell + month
  df %>%
    left_join(
      results %>% rename(!!month_col := mon),
      by = c("cell", month_col)
    ) %>%
    rename(!!new_var_colname := value)
}

addMonthFromYday <- function(df, yday_col = "onset") {
  df %>%
    dplyr::mutate(
      month = as.integer(format(
        as.Date(!!sym(yday_col), origin = "1970-01-01"), "%m"
      ))
    )
}

d_sp_cell <- readRDS("d_sp_cell.rds")
d_sp_cell <- addMonthFromYday(d_sp_cell,yday_col="onset")

d_sp_cell2 <- bindPrismByMonth(
  df              = d_sp_cell,
  prism_dir       = "D:/prism",
  month_col       = "month",        # your column name
  years           = 2021:2023,
  type            = "tmean",
  new_var_colname = "prism_tmean"
)

saveRDS(d_sp_cell2, "d_sp_cell2.rds")
```

Prep d_sp_cell2 for models

```{r}

d_sp_cell$vp_sc     <- scale(d_sp_cell$vp)[, 1]
d_sp_cell$tmean_sc  <- scale(d_sp_cell$tmean)[, 1]
d_sp_cell$srad_sc   <- scale(d_sp_cell$srad)[, 1]
d_sp_cell$onset_sc  <- scale(d_sp_cell$onset_corrected)[, 1]
d_sp_cell$cell_lat_sc  <- scale(d_sp_cell$cell_lat)[, 1]

d_sp_cell <- dplyr::filter(d_sp_cell, onset < 220)
```

Brms models

```{r}

#hist(d_sp_cell$black_30)
d_sp_cell2 <- d_sp_cell
#d_sp_cell2 <- readRDS("d_sp_cell2.rds")
d_sp_cell2$black <- d_sp_cell2$black_20
#d$black[d$black == 1] <- 0.9999
#d$black[d$black == 0] <- 0.0001

#d_sp_cell2$tmean_in_cell_onset <- scale(d_sp_cell2$prism_tmean)
d_sp_cell2$cell_lat <- scale(d_sp_cell2$cell_lat)
d_sp_cell2$onset <- scale(d_sp_cell2$onset)
d_sp_cell2$cell <- as.factor(d_sp_cell2$cell)
d_sp_cell2$species <- as.factor(d_sp_cell2$species)
d_sp_cell2 <- as.data.frame(d_sp_cell2)
d_sp_cell2$cell_lat <- d_sp_cell2$cell_lat[,1]
d_sp_cell2$onset <- d_sp_cell2$onset[,1]
d_sp_cell2$tmean <- scale(d_sp_cell2$tmean)
d_sp_cell2$tmean <- d_sp_cell2$tmean[,1]

formula <- "black ~ onset * cell_lat + (1|species) + (1|cell)" #+ tmean_in_cell_onset 
hist(d_sp_cell2$black)
max(d_sp_cell2$onset)

d <- d_sp_cell2
d <- as.data.frame(d)
d <- na.omit(d)
data2 <- d_sp_cell2 #make this a tree later
myBRMS <- function(formula){
  library(brms)
  m <- brm(
        formula = formula,
        data = d, #_filtered #for_tree
        data2 = data2,
        chains = 3,
        cores = 3,
        iter = 3000,
        warmup = 1000,
        control = list(max_treedepth = 15),
        #family = zero_inflated_beta(), 
        #family = gaussian(), 
    )
  m <- add_criterion(m,  criterion="loo")
  return(m)
}

library(sjPlot)
m <- myBRMS(formula)

plot_model(m,title = "mean lightness")
plot_model(m,terms="cell_lat",type="pred",
           axis.title = c("Scaled 10th Weib Percentile Onset Yday", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms="tmean_in_cell_onset",type="pred",
           axis.title = c("Tmean in cell onset", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms=c("onset","cell_lat"),type="int")

summary(m)
sjPlot::plot_model(m, type = "re")
# install.packages("performance")
library(performance)
m <- readRDS("m.rds")
saveRDS(m,file="m.rds")

library(lme4)

m2 <- lmer(black ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell), data=d_sp_cell2, na.action=na.omit)

library(car)
summary(m2)
vif(m2)
summary(m2)
plot_model(m2)
plot_model(m2,terms="cell_lat",type="pred")
summary(m2)
# returns a tibble with R2_marginal and R2_conditional
performance::r2(m)

plot_model(
  m2,
  type = "pred",
  terms = c("cell_lat [-2, 0, 2]","onset")# Replace with appropriate scaled values
)

myLMER <- function(formula){
  
}
d_sp_cell2 <- as.data.frame(d_sp_cell2)
d_sp_cell2$cell_lat <- d_sp_cell2$cell_lat[,1]
d_sp_cell2$onset <- d_sp_cell2$onset[,1]
typeof(d_sp_cell2)#$cell_lat
hist(d_sp_cell2$black)

```

LMER version

```{r}
library(sjPlot)

f1 <- "black_20 ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell)"
f2 <- "black_30 ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell)"
f3 <- "mean_lightness ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell)"

f4 <- "black_20 ~ onset * cell_lat + (1|species) + (1|cell)"
f5 <- "black_20 ~ onset + cell_lat + (1|species) + (1|cell)"

f6 <- "black_30 ~ onset * cell_lat + (1|species) + (1|cell)"
f7 <- "black_30 ~ onset + cell_lat + (1|species) + (1|cell)"

f8 <- "mean_lightness ~ onset * cell_lat + (1|species) + (1|cell)"
f9 <- "mean_lightness ~ onset + cell_lat + (1|species) + (1|cell)"

f10 <- "black_20 ~ onset * splines::ns(cell_lat,2) + wing_length + flight_type + (1|species) + (1|cell)"


#f11 <- "black_20 ~ onset * splines::ns(cell_lat,2) + ns(vpd,2) + (1|species) + (1|cell)"
f11 <- "black_20 ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell)"


f12 <- "black_20 ~ onset * splines::ns(vpd,2) + (1|species) + (1|cell)"

f13 <- "black_20 ~ onset * splines::ns(tmean,2) + (1|species) + (1|cell)"

f14 <- "black_20 ~ onset * splines::ns(vpd_max,2) + (1|species) + (1|cell)"
f15 <- "black_20 ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell)"

f16 <- "black_20 ~ onset * splines::ns(tmean, 2) + onset * splines::ns(vpd_max, 2) + (1 | species) + (1 | cell)"

f16 <- "black_20 ~ (onset + splines::ns(tmean, 2) + splines::ns(vpd_max, 2))^2 + (1 | species) + (1 | cell)"

f17 <- "black_20 ~ (onset + splines::ns(tmean, 2) + splines::ns(vpd_max, 2))^2 + (1 | species) + (1 | cell)"


view(d_sp_cell2)
plot_model(m11, type="pred",terms="vpd")
d <- d_sp_cell2
data2 <- d_sp_cell2 #make this a tree later
myLMER <- function(formula, plot=TRUE){
  library(lme4)
  m <- lmer(
        formula = formula,
        data = d
    )
  if(!plot){
    return(m)
  }
  plot(plot_model(m))
  plot(plot_model(m,terms="cell_lat",type="pred"))
  plot(plot_model(m, type = "pred", terms = c("cell_lat [-2, 0, 2]","onset")))
  print(performance::r2(m))
  print(AIC(m))
  return(m)
}

plot_model(m11, type = "pred", terms = c("cell_lat [-2, 0, 2]","vpd"))
plot_model(m11)
m1 <- myLMER(f1)
m2 <- myLMER(f2)
m3 <- myLMER(f3)

m4 <- myLMER(f4)
m5 <- myLMER(f5)

m6 <- myLMER(f6)
m7 <- myLMER(f7)

m8 <- myLMER(f8)
m9 <- myLMER(f9)

m10 <- myLMER(f10)

library(splines)
m11 <- myLMER(f11,plot=FALSE)

m12 <- myLMER(f12,plot=FALSE)
m13 <- myLMER(f13,plot=FALSE)
m14 <- myLMER(f14,plot=FALSE)
m15 <- myLMER(f15,plot=FALSE)
m16 <- myLMER(f16,plot=FALSE)

AIC(m11)
AIC(m12)
AIC(m13)
AIC(m14)
AIC(m15)
AIC(m16)


aic(m13)
plot_model(m14, type = "pred", terms = c("vpd_max","onset [-2, 0, 2]"))
plot_model(m13, type = "pred", terms = c("tmean","onset [-2, 0, 2]"))
library(sjPlot)
plot_model(m14, type = "pred", terms = "vpd_max")
plot_model(m13, type = "pred", terms = "tmean")

summary(m11)

myBRMS <- function(formula){
  library(brms)
  m <- brm(
        formula = formula,
        data = d, #_filtered #for_tree
        data2 = data2,
        chains = 3,
        cores = 3,
        iter = 3000,
        warmup = 1000
        #family = zero_inflated_beta(), 
        #family = gaussian(), 
    )
  m <- add_criterion(m,  criterion="loo")
  return(m)
}

m1 <- myBRMS(f1)
saveRDS(m1,"m1.rds")
m2 <- myBRMS(f2)
saveRDS(m2,"m2.rds")
m3 <- myBRMS(f3)
saveRDS(m3,"m3.rds")
m4 <- myBRMS(f4)
saveRDS(m4,"m4.rds")
m5 <- myBRMS(f5)
saveRDS(m5,"m5.rds")
m6 <- myBRMS(f6)
saveRDS(m6,"m6.rds")
m7 <- myBRMS(f7)
saveRDS(m7,"m7.rds")
m8 <- myBRMS(f8)
saveRDS(m8,"m8.rds")
m9 <- myBRMS(f9)

m11 <- myBRMS(f11)
summary(m11)

library(sjPlot)
plot_model(m1)
plot_model(m1,terms="cell_lat",type="pred",show.data=TRUE)
plot_model(m1, type = "pred", terms = c("cell_lat [-2, 0, 2]","onset"),show.data=TRUE,alpha=0.2)

plot_model(m2)
plot_model(m2,terms="cell_lat",type="pred",show.data=TRUE)
plot_model(m2, type = "pred", terms = c("cell_lat [-2, 0, 2]","onset"),show.data=TRUE)

loo_compare(m2,m6,m7)
loo_compare(m3,m8,m9)

summary(m2)
plot_model(m)
plot_model(m,terms="cell_lat",type="pred")
summary(m2)
performance::r2(m)

plot_model(
  m2,
  type = "pred",
  terms = c("cell_lat [-2, 0, 2]","onset")# Replace with appropriate scaled values
)


loo_compare(m1,m4,m5)

library(sjPlot)
m <- myBRMS(formula)

plot_model(m,title = "mean lightness")
plot_model(m,terms="cell_lat",type="pred",
           axis.title = c("Scaled 10th Weib Percentile Onset Yday", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms="tmean_in_cell_onset",type="pred",
           axis.title = c("Tmean in cell onset", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms=c("onset","cell_lat"),type="int")

summary(m)
sjPlot::plot_model(m, type = "re")
# install.packages("performance")
library(performance)
m <- readRDS("m.rds")
saveRDS(m,file="m.rds")

library(lme4)

m2 <- lmer(black ~ onset * splines::ns(cell_lat,2) + (1|species) + (1|cell), data=d_sp_cell2, na.action=na.omit)

plot(plot_model(m11, type = "pred", terms = c("cell_lat","vpd")))

library(car)
summary(m2)
vif(m2)
summary(m2)
plot_model(m1)
plot_model(m2,terms="cell_lat",type="pred",show.data=TRUE)
summary(m2)
# returns a tibble with R2_marginal and R2_conditional
performance::r2(m)


plot_model(m1, type = "pred", terms = c("onset","cell_lat [-2, 0, 2]"),show.data=TRUE,alpha=0.1)

plot_model(m10, type = "pred", terms = "flight_type",show.data=FALSE)

myLMER <- function(formula){
  
}

car::vif(m11)
cor.test(d_sp_cell2$vpd,d_sp_cell2$cell_lat,na.action="omit")

d_sp_cell2 <- as.data.frame(d_sp_cell2)
d_sp_cell2$cell_lat <- d_sp_cell2$cell_lat[,1]
d_sp_cell2$onset <- d_sp_cell2$onset[,1]
typeof(d_sp_cell2)#$cell_lat
hist(d_sp_cell2$black)
summary(m11)
car::vif(m11)

plot_model(m13,type="pred")

plot_model(m13, type = "pred", terms = c("tmean [-2, 0, 2]","onset"),show.data=TRUE,alpha=0.1)

car::vif(m16)

plot_model(m16)
summary(m16)
# lat model is way better than either tmean and/or vpd 

AIC(m11)
AIC(m12)
AIC(m16)
```

```{r}

m_lat <- lmer(black_20 ~ onset + splines::ns(cell_lat, 2) + (1|species) + (1|cell), data = d)
m_vpd <- lmer(black_20 ~ onset + splines::ns(vpd, 2) + (1|species) + (1|cell), data = d)
AIC(m_lat)
AIC(m_vpd)
```

```{r}

p <- plot_model(m1, type = "pred", terms = c("onset","cell_lat [-2, 0, 2]"),show.data=FALSE,alpha=0.1)

d3 <- dplyr::filter(d_sp_cell2,cell_lat>=1)

m <- lm(black_20 ~ onset, data=d3)
plot_model(m)
hist(d_sp_cell2$cell_lat)
p + 
  geom_point(aes(x = onset, y = black_20), 
             data = d3,  # this is where raw data is stored
             color = "black", 
             alpha = 0.05,
             inherit.aes = FALSE)
```

```{r}

library(sjPlot)
library(ggplot2)

p <- plot_model(
  m1,
  type = "pred",
  terms = c("cell_lat [-2, 0, 2]", "onset"),
  show.data = FALSE,
  alpha = 0.1
)
p
# Remove the default data points and re-add them manually in black
# Unfortunately, sjPlot doesn't provide a clean way to extract the raw data,
# so we just layer black points on top with the right aesthetics:

p + 
  geom_point(aes(x = cell_lat, y = black_20), 
             data = d,  # this is where raw data is stored
             color = "black", 
             alpha = 0.05,
             inherit.aes = FALSE)

p <- plot_model(
  m1,
  type = "pred",
  terms = c("onset", "cell_lat [-2, 0, 2]"),
  show.data = FALSE,
  alpha = 0.1
)
p
# Remove the default data points and re-add them manually in black
# Unfortunately, sjPlot doesn't provide a clean way to extract the raw data,
# so we just layer black points on top with the right aesthetics:

p + 
  geom_point(aes(x = onset, y = black_20), 
             data = d,  # this is where raw data is stored
             color = "black", 
             alpha = 0.05,
             inherit.aes = FALSE)
```

```{r}

d_sp_cell2 <- readRDS("d_sp_cell2.rds")
d_sp_cell2$black <- d_sp_cell2$black_20
d_sp_cell2$black <- scale(d_sp_cell2$black)
#d$black[d$black == 1] <- 0.9999
#d$black[d$black == 0] <- 0.0001
d_sp_cell2$tmean_in_cell_onset <- scale(d_sp_cell2$prism_tmean)
d_sp_cell2$cell_lat <- scale(d_sp_cell2$cell_lat)
d_sp_cell2 <- dplyr::filter(d_sp_cell2, onset < 220)
d_sp_cell2$onset <- scale(d_sp_cell2$onset)
d_sp_cell2$cell <- as.factor(d_sp_cell2$cell)
d_sp_cell2$species <- as.factor(d_sp_cell2$species)
formula <- "black ~ onset * cell_lat + tmean_in_cell_onset + (1|species) + (1|cell)" #+ tmean_in_cell_onset 
hist(d_sp_cell2$black)
max(d_sp_cell2$onset)

d <- d_sp_cell2
d <- as.data.frame(d)
data2 <- d_sp_cell2 #make this a tree later
myBRMS <- function(formula){
  library(brms)
  m <- brm(
        formula = formula,
        data = d, #_filtered #for_tree
        data2 = data2,
        chains = 3,
        cores = 3,
        iter = 3000,
        warmup = 1000,
        control = list(max_treedepth = 15),
        #family = zero_inflated_beta(), 
        #family = gaussian(), 
    )
  m <- add_criterion(m,  criterion="loo")
  return(m)
}

library(sjPlot)
m <- myBRMS(formula)

plot_model(m,title = "mean lightness")
plot_model(m,terms="cell_lat",type="pred",
           axis.title = c("Scaled 10th Weib Percentile Onset Yday", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms="tmean_in_cell_onset",type="pred",
           axis.title = c("Tmean in cell onset", "Scaled Mean Lightness"),
           title = "")
plot_model(m,terms=c("onset","cell_lat"),type="int")

summary(m)
sjPlot::plot_model(m, type = "re")
# install.packages("performance")
library(performance)

# returns a tibble with R2_marginal and R2_conditional
performance::r2(m)
```

Interaction plotting

```{r}

lat_vals <- c(-2, 0, 2)  
# how many points along 'onset'?
n_onset <- 100  

newdata <- expand.grid(
  onset    = seq(min(d$onset), max(d$onset), length.out = n_onset),
  cell_lat = lat_vals,
  # hold the other predictor at its mean:
  tmean_in_cell_onset = mean(d$tmean_in_cell_onset, na.rm = TRUE),
  # for group‐level factors we set NA (so we only get population‐level fit):
  species  = NA,  
  cell     = NA   
)
# 'fitted()' returns a matrix with columns Estimate, Q2.5, Q97.5, etc.
pred_mat <- fitted(m, newdata = newdata, re_formula = NA)

# tack them onto newdata
newdata$fit    <- pred_mat[, "Estimate"]
newdata$lower  <- pred_mat[, "Q2.5"]
newdata$upper  <- pred_mat[, "Q97.5"]

library(ggplot2)

ggplot(newdata, aes(x = onset, y = fit, color = factor(cell_lat))) +
  # main lines
  geom_line(size = 1) +
  # ribbon for credible interval
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = factor(cell_lat)),
              alpha = 0.2, color = NA) +
  scale_color_brewer("Latitude\n(std)", palette = "Dark2") +
  scale_fill_brewer("Latitude\n(std)", palette = "Dark2") +
  labs(
    x = "Scaled Onset (|day – 172|, standardized)",
    y = "Predicted Black",
    title = "How the effect of Onset varies across Latitude"
  ) +
  theme_minimal()


```

Failed attempt to bind daily data

```{r}

d_sp_cell <- readRDS("d_sp_cell.rds")

library(purrr)
bindPrismByYdayOnOff <- function(df,
                                 prism_dir,
                                 cell_width_km   = 250,
                                 variable        = "tmean",
                                 avg_of_years    = 2017:2023,
                                 prefix          = "prism_tmean") {
  #— libs
  library(dplyr);  library(geosphere)
  library(raster); library(prism)
  library(purrr);  library(colorEvoHelpers)
  
  #— point PRISM at your download dir
  prism_set_dl_dir(prism_dir)
  
  #— make one polygon per unique cell
  uniq <- df %>%
    dplyr::select(cell, yday_onset, yday_offset, cell_lat, cell_lon) %>%
    distinct() %>%
    rowwise() %>%
    mutate(
      cell_poly = list(
        cellPolyFromLatLonWidth(cell_lat, cell_lon, cell_width_km)
      )
    ) %>%
    ungroup()
  
  #— helper: pull daily rasters, average, extract mean for a polygon
  getDayStats <- function(years, yday, cell_poly, variable) {
    if (inherits(cell_poly, "sf")) {
      cell_poly <- as(cell_poly, "Spatial")
    }

    dates    <- as.Date(yday - 1, origin = paste0(years, "-01-01"))
    date_str <- format(dates, "%Y-%m-%d")

    pd    <- prism_archive_subset(variable, "daily", dates = date_str)
    files <- pd_to_file(pd)

    rasters    <- lapply(files, raster)
    raster_avg <- calc(stack(rasters), fun = mean)

    r_crop <- crop(raster_avg, extent(cell_poly))
    r_mask <- mask(r_crop, cell_poly)

    cellStats(r_mask, stat = "mean")
  }

  #— compute onset and offset stats
  uniq <- uniq %>%
    mutate(
      !!paste0(prefix, "_onset") := pmap_dbl(
        list(avg_of_years, yday_onset, cell_poly),
        getDayStats,
        variable = variable
      ),
      !!paste0(prefix, "_offset") := pmap_dbl(
        list(avg_of_years, yday_offset, cell_poly),
        getDayStats,
        variable = variable
      )
    )

  #— join back to original df
  out <- df %>%
    left_join(
      uniq %>% dplyr::select(cell, yday_onset, yday_offset,
                      starts_with(prefix)),
      by = c("cell", "yday_onset", "yday_offset")
    )

  return(out)
}


d_sp_cell$yday_onset <- d_sp_cell$onset
d_sp_cell$yday_offset <- d_sp_cell$offset

test <- bindPrismByYdayOnOff(
  df           = d_sp_cell,
  prism_dir    = "D:/prism",
  cell_width_km= 250,
  variable     = "tmean",
  avg_of_years = 2021:2023,       # average climatology over these years
  prefix       = "prism_tmean"    # you'll get prism_tmean_onset & prism_tmean_offset
)
```
